---
title: Computing RNA velocity in a Bioconductor framework
author:
- name: Kevin "the Destroyer" Rue-Albrecht
- name: Aaron "Cholesterol" Lun
  email: infinite.monkeys.with.keyboards@gmail.com
- name: Charlotte "Sledgehammer" Soneson
package: velociraptor
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{User's guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
library(BiocStyle)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Overview

This package provides a lightweight interface between the Bioconductor `SingleCellExperiment` data structure
and the `r basilisk::PyPiLink("scvelo")` Python package for RNA velocity calculations.
The interface is comparable to that of many other `SingleCellExperiment`-compatible functions,
allowing users to plug in RNA velocity calculations into the existing Bioconductor analysis framework.
To demonstrate, we will use some spliced/unspliced data from the `r Biocpkg("scRNAseq")` package.

```{r}
library(scRNAseq)
sce <- HermannSpermatogenesisData()
sce
```

# Basic workflow

We assume that feature selection has already been performed by the user using any method 
(see [here](https://osca.bioconductor.org/feature-selection.html) for some suggestions).
We will use the log-expression method from the `r Biocpkg("scran")` method here.

```{r}
library(scuttle)
sce <- logNormCounts(sce, assay.type=1)

library(scran)
dec <- modelGeneVar(sce)
top.hvgs <- getTopHVGs(dec, n=2000)
```

We can plug these choices into the `scvelo()` function with our `SingleCellExperiment` object.
Here, we will use the `"spliced"` count matrix as a proxy for the typical exonic count matrix.
There are some subtle differences between the two - see `?scvelo` for some commentary about this -
but the spliced counts are generally a satisfactory replacement.

```{r}
library(velociraptor)
velo.out <- scvelo(sce, subset.row=top.hvgs, assay.X="spliced")
velo.out
```

`scvelo()` produces a `SingleCellExperiment` containing all of the outputs of the calculation in Python.
Of particular interest is the `velocity_pseudotime` vector that captures the relative progression of each cell 
along the biological process driving the velocity vectors.
We can visualize this effect below in a $t$-SNE plot generated by `r Biocpkg('scater')` on the top HVGs.

```{r}
library(scater)

set.seed(100)
sce <- runPCA(sce, subset_row=top.hvgs)
sce <- runTSNE(sce, dimred="PCA")

sce$velocity_pseudotime <- velo.out$velocity_pseudotime
plotTSNE(sce, colour_by="velocity_pseudotime")
```

And that's it, really.

# Advanced options

`scvelo()` interally performs a PCA step that we can bypass by supplying our own PC coordinates.
Indeed, it is often the case that we have already performed PCA in the earlier analysis steps,
so we can just re-use those results to (i) save time and (ii) improve consistency with the other steps.
Here, we computed the PCA coordinates in `runPCA()` above, so let's just recycle that:

```{r}
# Only setting assay.X= for the initial AnnData creation,
# it is not actually used in any further steps.
velo.out2 <- scvelo(sce, assay.X=1, subset.row=top.hvgs, use.dimred="PCA") 
velo.out2
```

We also provide an option to use the `r basilisk::PyPiLink("scvelo")` pipeline without modification,
i.e., relying on their normalization and feature selection.
This sacrifices consistency with other Bioconductor workflows but enables perfect mimicry of a pure Python-based analysis. 
In this case, arguments like `subset.row=` are simply ignored. 

```{r}
velo.out3 <- scvelo(sce, assay.X=1, use.theirs=TRUE)
velo.out3
```

Advanced users can tinker with the settings of individual `r basilisk::PyPiLink("scvelo")` steps
by setting named lists of arguments in the `scvelo.params=` argument.
For example, to tinker with the behavior of the `recover_dynamics` step, we could do:

```{r}
velo.out4 <- scvelo(sce, assay.X=1, subset.row=top.hvgs,
    scvelo.params=list(recover_dynamics=list(max_iter=20)))
velo.out4
```

# Session information

```{r}
sessionInfo()
```
